---
title: "Homework 1"
author: "Ashutosh Ekade"
date: "Due Oct 9, 2023"
format: pdf
---

## Question 1: Expectations

#### 1.a) Define both and explain the difference between (1) the expectation of a random variable and (2) the sample average.

Answer:

a) The expectation of a random variable is a measure of the central tendency of the random variable's distribution and is the average value with probability weighted averaging. It is defined as:

$\mathbb{E}[g(X)]=\sum_{i} g(X_i)p_x(x_i)$ for discrete random variable x if the series is convergent, and,

$\mathbb{E}[g(X)]=\int_{-\infty}^{\infty} g(X_i)f_x(x_i)dx$ for continuous random variable x.

b) The Sample Average

Sample average or the sample mean of a random draw from a larger population of X is defined as:

$\bar{X}_n = \frac{1}{n}\sum_{i}^{n} X_i$ where n is the sample size.

Difference:

Sample average is a function of the sample drawn from the population whereas the expected value is not. Expected value is a weighted average whereas sample average is not.


## Question 2: LLN & CLT

#### 2.a) Plot the density of a Beta(5,2) distribution over it's domain [0,1]. Make a publication-quality plot by changing any unwanted default plotting behavior and by adding relevant titles and labels. The first parameter of the Beta distribution is often labeled $\alpha$ and the second $\beta$. The `dbeta()` and related functions in R label these parameters as `shape1` and `shape2`.

```{r}
# ...add answer here...
x <- runif(500) # generates 500 uniformly distributed random numbers between 0-1 (default range)
y <- dbeta(x, 5, 2)
plot(x=x, y=y, xlab="x", ylab="Beta(x, 5, 2)", main="Density of Beta(alpha, beta) distribution",
     col = "black", pch = 20, cex = 0.5,
     xlim = c(0, 1), 
     ylim = c(0, 2.5), 
     yaxs="i", xaxs="i",
     sub = "alpha=5, beta=2")

```

#### 2.b) State the Law of Large Numbers as simply as you can.

The law of large numbers states that if we repeat an experiment a large number of times and take the average of all the results then this average value approaches the expected value of the experiment.

#### 2.c) Set the seed to the value 1234 (`set.seed(1234)`). Then take 1,000 random draws from the Beta(5,2) distribution using `rbeta()`. Calculate a running sample average. Specifically: calculate $\bar{X}_1 = x_1$, then calculate $\bar{X}_2 = (1/2) \sum_{i=1}^2 X_i$, then calculate $\bar{X}_3 = (1/3) \sum_{i=1}^3 X_i$. Continue until you have calculated $\bar{X}_{1000} = (1/1000) \sum_{i=1}^{1000} X_i$. Create a scatterplot with the values 1--1,000 on the horizontal axis and the 1,000 cumulative average values of $\bar{X}_i$ for $i=1,\ldots,1000$ you calculated on the vertical axis. Compare your value for $\bar{X}_{1000}$ to the $\mathbb{E}[X]=\alpha/(\alpha+\beta)=5/7 = 0.7143$.

```{r}
set.seed(1234)
y <- vector(mode="list", length = 1000)
x <- rbeta(1000, 5, 2)
for (i in 1:1000) {
  y[i] <- mean(x[1:i])
}
plot(x=1:1000, y=y, xlab="i", ylab="X̄", main="Demonstration of the Law of Large Numbers",
     col = "black", pch = 20, cex = 0.5,
     xlim = c(0, 1000), 
     ylim = c(0.65, 0.8), 
     yaxs="i", xaxs="i")
```
```{r}
print(5 / 7)
```
```{r}
print(y[1000])
```
#### 2.d) State the Central Limit Theorem as simply as you can.

The Central limit Theorem states that if we draw multiple large samples with replacement from a population then the sample mean will follow a normal distribution irrespective of the distribution of the population as long as the sample we draw is sufficiently large.

#### 2.e) Make two plots. For the first plot, take D=10 draws from the Beta(5,2) distribution and calculate the sample average. Repeat the process of taking D=10 draws and finding the sample average R=10,000 times. Plot a histogram of the 10,000 sample averages. For the second plot, repeat the process with D=100 draws. These two histograms are called "sampling distributions." 

```{r}
draws <- 10
iterations <- 10000
y <- numeric(length=iterations)
for (i in 1:iterations){
  x <- rbeta(draws, 5, 2)
  y[i] <- mean(x)
}
hist(y, main = "Sampling Distribution 1", xlab = "Sample Average", ylab = "Frequency",
     xlim = c(0.5, 1), 
     ylim = c(0, 2000), 
     yaxs="i", xaxs="i")
```

```{r}
draws <- 100
iterations <- 10000
y <- numeric(length=iterations)
for (i in 1:iterations){
  x <- rbeta(draws, 5, 2)
  y[i] <- mean(x)
}
hist(y, main = "Sampling Distribution 2", xlab = "Sample Average", ylab = "Frequency",
     xlim = c(0.5, 1), 
     ylim = c(0, 3000), 
     yaxs="i", xaxs="i")
```
## Question 3: Linear Algebra

**Suppose $\mathbf{X}$ and $\mathbf{Y}$ are defined as follows.**

$$ \mathbf{X} = \begin{bmatrix} 1 & 1 \\ 1 & 4 \\ 1 & 5 \\ 1 & 8 \end{bmatrix}  \hspace{5em} \mathbf{y} = \begin{bmatrix} 6 \\ 5 \\ 3 \\ 2 \end{bmatrix} $$

#### 3.a) What is the rank of $\mathbf{X}$? Provide a brief (approx 1 sentence) explanation. Check your work via R code using `Matrix::rankMatrix()`.

Rank of Matrix X is 2. Rank of a matrix is the highest number of independent rows or columns in a matrix. If we see in matrix X, row 3 can be created by adding row 1 & row 2, whereas row 4 is 2 times row 1. So we have 2 independent rows in X. We can verify that too in the code below.

```{r}
library(Matrix)
X <- matrix(c(1, 1, 1, 1, 1, 4, 5, 8), 4, 2); X
t(X)
y <- matrix(c(6, 5, 3, 2), 4, 1); y
r <- rankMatrix(X)
print(r[1])
```

#### 3.b) Calculate $\mathbf{X}'\mathbf{X}$. Use the `bmatrix` environment in Latex to typeset your answer. Check your work via R code. 

Matrix $\mathbf{X}$ = $\begin{bmatrix} 1 & 1 \\ 1 & 4 \\ 1 & 5 \\ 1 & 8 \\ \end{bmatrix}$
Matrix $\mathbf{X}'$ = $\begin{bmatrix} 1 & 1 & 1 & 1 \\ 1 & 4 & 5 & 8 \\ \end{bmatrix}$
$\mathbf{X}'\mathbf{X}$ = $\begin{bmatrix} 1 & 1 & 1 & 1 \\ 1 & 4 & 5 & 8 \\ \end{bmatrix}$ $\begin{bmatrix} 1 & 1 \\ 1 & 4 \\ 1 & 5 \\ 1 & 8 \\ \end{bmatrix}$ = $\begin{bmatrix} 4 & 18 \\ 18 & 106 \\ \end{bmatrix}$

```{r}
t(X)%*%X
```

#### 3.c) What is the rank of $\mathbf{X}'\mathbf{X}$? Provide a brief explanation. Check your work via R code.

As we can see, in matrix $\mathbf{X}'\mathbf{X}$, both the rows (or columns) are independent and one cannot be expressed as a multiple of the other. Hence the rank of $\mathbf{X}'\mathbf{X}$ is 2.

$\begin{bmatrix} 1 & 1 & 1 & 1 \\ 1 & 4 & 5 & 8 \\ \end{bmatrix}$ $\begin{bmatrix} 1 & 1 \\ 1 & 4 \\ 1 & 5 \\ 1 & 8 \\ \end{bmatrix}$

*Snippet for LaTex bmatrix is used with the help of ChatGPT.
```{r}
r <- rankMatrix(t(X)%*%X)
print(r[1])
```

#### 3.d) Find $(\mathbf{X}'\mathbf{X})^{-1}$ "by hand" (as you would with paper and pencil) using the approach outlined on slide 47 of the Class 1 slides. Check your work via R code.

$\mathbf{X}'\mathbf{X}$ = $\begin{bmatrix} 4 & 18 \\ 18 & 106 \\ \end{bmatrix}$

$\mathbf{X}'\mathbf{X}^{-1}$ = $\frac{1}{|\mathbf{X}'\mathbf{X}|}$ $\begin{bmatrix} 106 & -18 \\ -18 & 4 \\ \end{bmatrix}$
 = $\frac{1}{(4*106 - 18*18)}$ $\begin{bmatrix} 106 & -18 \\ -18 & 4 \\ \end{bmatrix}$
 = $\frac{1}{(100)}$ $\begin{bmatrix} 106 & -18 \\ -18 & 4 \\ \end{bmatrix}$
 = $\begin{bmatrix} 1.06 & -0.18 \\ -0.18 & 0.04 \\ \end{bmatrix}$

```{r}
z <- t(X)%*%X
d <- solve(z); d
```

#### 3.e) What is the rank of $(\mathbf{X}'\mathbf{X})^{-1}$? Provide a brief explanation. Check your work via R code.

All rows (or columns) of matrix $(\mathbf{X}'\mathbf{X})^{-1}$ are independent, so its rank is 2.

```{r}
rankMatrix(d)[1]
```

#### 3.f) Calculate $\mathbf{X}'\mathbf{y}$. Check your work via R code.

 = $\begin{bmatrix} 1 & 1 & 1 & 1 \\ 1 & 4 & 5 & 8 \\ \end{bmatrix}$ $\begin{bmatrix} 6 \\ 5 \\ 3 \\ 2 \\ \end{bmatrix}$
 = $\begin{bmatrix} (1*6 + 1*5 + 1*3 + 1*2) \\ (1*6 + 4*5 + 5*3 + 8*2) \\ \end{bmatrix}$
 = $\begin{bmatrix} 16 \\ 57 \\ \end{bmatrix}$

```{r}
f <- t(X) %*% y; f
```

#### 3.g) Use your results from 3d and 3f to calculate "by hand" $(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{y}$.  Check your work via R code.

= $\begin{bmatrix} 4 & 18 \\ 18 & 106 \\ \end{bmatrix}$ $\begin{bmatrix} 16 \\ 57 \\ \end{bmatrix}$
= $\begin{bmatrix} (4*16 + 18*57) \\ (18*16 + 106*57) \\ \end{bmatrix}$
= $\begin{bmatrix} 1090 \\ 6330 \\ \end{bmatrix}$

```{r}
g <- z %*% f; g
```




